Cross-relaxation in dense ensembles of NV centers and application to magnetometry

### 2022-10-19 Manuscrit Thèse/these_fusion.tex
toto

\documentclass[a4paper,11pt]{report}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric-comp,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{bib_thesis.bib} %Import the bibliography file

@@

\DeclareUnicodeCharacter{0308}{HERE!HERE!} % je vais pas m'énerver, mais un peu quand meme

\title{}
\begin{document}
\begin{refsection}

\chapter{Introduction to Nitrogen-Vacancy centers and their common usage}
The aim of this first chapter is to introduce concepts related to Nitrogen-vacancy (NV) centers, to the diamond material, and to the dipole-dipole interaction which are necessary to explain the work presented in the later chapters.

@@

\subsection{Experimental setup}
\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{chapter 1/Figures/shema_exp}
\caption{Confocal microscopy setup detailed in main text}
\label{setup}
\end{figure}
The three most common experimental degrees of freedom to consider for NV experiments are: the optical illumination, the microwave field and the external magnetic field. Other parameters such as the crystal temperature, the electric field or the crystal stress can also be controlled for specific applications.

Fig. \ref{setup} shows a multi-purpose experimental setup which offers a control over these three degree of freedoms. Unless specified otherwise, this is the experimental setup that was used to perform all the measurements shown in this manuscript.

@@

\subsection{NV spin $T_2$}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/Echo}
\caption{Spin-echo protocol. (a) Pulse seqeunce used. (b) Experimental data obtained on sample CVD-pink, fitted with an exponential decay of time $3.9\ \rm{\mu s}$.} 
\label{Echo}
\end{figure}
\subsubsection{Measurement protocol}
As precised earlier, the exact $T_2$ definition depends on the rephasing scheme used. We will focus here on the simplest known rephasing process which is the spin-echo or Hahn-echo sequence.

@@

\subsection{Cross-relaxations with NV centers}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/Shema_CR}
\caption{Illustration of cross-relaxations between an NV center and an unpolarized spin. Red dots represent the population in each states, and the arrows the various population transfer mechanisms}
\label{CR_shema}
\end{figure}

@@

The focus of chapter 2 will be on the cross-relaxation between NV centers and other spin impurities in CVD-grown diamond, while the focus of chapter 3 and 4 will be on the cross-relaxation between NV centers and other NV centers, as well as their potential application in magnetometry.
\printbibliography
\end{refsection}
 \begin{refsection}

\chapter{Detection of dark spins via cross-relaxations}
In this chapter, we will see how the interplay between the optical and spin properties of the NV centers can be transferred to other spins in the crystal by coupling them to NV centers. More specifically, I will discuss the first optical detection of two spin species, VH$^-$ and War1, in a CVD diamond by the mean of cross-relaxations with NV centers.
\section{Flip-Flops, double flips and cross-relaxation}
\label{Sec_CR}

\subsection{Resonant interaction : flip-flops and double flips}

The dominant form of interaction between two distant spins is the magnetic dipole-dipole interaction. For two spins with with vector spin Hamiltonian $\hat{\mathbf{S}}_1$ and $\hat{\mathbf{S}}_2$, separated by a distance $\mathbf{r}=r\mathbf{u}$, the dipole-dipole interaction Hamiltonian reads (see Appendix A) :
\begin{equation}
\mathcal{H}_{dd}=\frac{J_0}{r^3}\left[ \hat{\mathbf{S}}_1\cdot\hat{\mathbf{S}}_2 - 3 (\hat{\mathbf{S}}_1\cdot\mathbf{u})(\hat{\mathbf{S}}_2\cdot\mathbf{u})\right],
\end{equation}
where $J_0= \frac{\mu_0\gamma_1\gamma_2 \hbar^2}{4\pi}$. For two electronic spins with $g-$factors close to 2 (which is the case for the NV center and most spin defect in diamond), the numerical value of $J_0$ is $(2\pi)\,52\ \rm{MHz}\cdot\rm{nm}^3$.

Flip-flop, in the context of dipole-dipole interaction, is the mechanism which exchanges a quantum of spin between two spins. For an initial state of two spins $\ket{m_s^1=i;m_s^2=j}$, a flip-flop process will result in a final state $\ket{i+1;j-1}$ or $\ket{i-1;j+1}$. The effective rate of this process depends on the matrix element $\mel{i;j}{\mathcal{H}_{dd}}{i\pm 1;j \mp 1}$, as well as the resonance condition between the initial and final state. We should note that for two identical spins, the flip-flop process $\ket{i;i\pm 1}\bra{i\pm 1,i}$ is always resonant.

Another mechanism authorized by the dipole-dipole Hamiltonian is the double-flip, either up or down, which couple the states $\ket{i;j}$ and $\ket{i-1;j-1}$ or $\ket{i+1;j+1}$. For a single spin species in a strong magnetic field, the lift of the various spin levels by the Zeeman effect means that no double-flip process can be resonant. However with weak magnetic field or when two different spin species are present, double flip process can be as important as flip-flops since the matrix elements $\mel{i;j}{\mathcal{H}_{dd}}{i\pm 1;j \pm 1}$ are typically of the same order of magnitude as the flip-flop ones.

\subsection{Cross-relaxations}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{chapter 2/Figures/Shema_CR}
\caption{Illustration of cross-relaxations between an NV center and an unpolarized spin. Red dots represent the population in each states, and the arrows the various population transfer mechanisms}
\label{CR_shema}
\end{figure}

Cross-relaxation (CR) is the transfer of polarization from one spin to another (or more generally from one family of spins to another family). This process con occur either through flip-flops, or through double flips, as long as the resonance condition between the two spins is met.

The reason why CR tend to depolarize the NV center simply comes from a rate equation : since the initial NV population is higher in the $\ket{0}$ state than in the $\ket{-1}$ state, the flip-flop (or double flip) process that makes the NV center go to $\ket{-1}$ (symbolized by the two red arrows in Fig. \ref{CR_shema}) is more likely than the reverse flip-flop. If we only focus on the NV center, cross-relaxation is simply another thermalization process due to the coupling of the NV center to its environment.

For a single NV center coupled to spin 2, assuming that spin 2 is initially depolarized and that the coupling strength between the two spins is smaller than their respective dephasing time $T_2^*$, we can compute the modification in the NV relaxation rate $\Gamma_1=\frac{1}{T_1}$ caused by spin 2 as \citep{hall2016detection}:
\begin{equation}
\delta \Gamma_1=\frac{\Omega^2}{\Gamma_2^*} \frac{(\Gamma_2^*)^2}{(\delta \nu)^2+(\Gamma_2^*)^2},
\label{delta gamma 1}
\end{equation}
where $\Omega=\frac{\mel{i;j}{\mathcal{H}_{\rm dd}}{j;i}}{\hbar}$ is the matrix element of the dipole-dipole Hamiltonian that couples the initial and final states, $\Gamma_2^*=\frac{1}{T_2^*(\rm NV)}+\frac{1}{T_2^*(\rm Spin 2)}$ is the joint dephasing rate of the NV center and the second spin, and $\delta \nu$ is the detuning between the central frequencies of the NV center and spin 2.
\section{The dark spins in diamond}

We refer by dark spins to spin defects which do not interact with light, which in the case of diamond defects consist of almost every known spin defect bar the NV center and the more recently found ST1 center \citep{lee2013readout, john2017bright}.

\subsection{P1 centers and $^{13}$C}
By far the most common spin defects in NV-rich diamond, outside of NV centers, are the substitutional nitrogen centers N$^0_S$ known as the P1 centers, which have an electronic spin $1/2$ and a nuclear spin 1, and the $^{13}$C $1/2$ nuclear spin which is present at 1.1\% in natural abundance carbon, the remaining carbon atoms being almost exclusively the spinless $^{12}$C isotope.

In NV-rich diamonds, these two impurities are the main causes of the inhomogeneous dephasing rate $\Gamma_2^*=1/T_2^*$ of the NV centers ensemble \citep{barry2020sensitivity}. The dephasing comes from the distribution of dipole-dipole coupling from each NV centers to its neighboring spin defects, which varies in space and time.

The dephasing rate caused by a given impurity is directly proportional to the diagonal dipole-dipole matrix elements \citep{bauch2020decoherence}: 
\begin{equation}
\Gamma_2^*(\rm P1) \propto \Omega_{\rm NV-P1}=\frac{1}{\hbar}\sqrt{\sum_j \abs{\mel{NV=i;P1=j}{\mathcal{H}_{\rm dd}}{NV=i;P1=j}}^2}.
\label{Gamma2 P1}
\end{equation}
Since the dipole-dipole coupling rate scales as $1/r^3$, and the distance $r$ between each spin scales as the inverse cubic root of the concentration,  it is expected that the dephasing rate associated with a particular impurity scales linearly with the concentration of the impurity: $\Gamma_2^*(\rm P1) \propto [P1]$. In the case of P1 centers, this was verified experimentally by \citep{bauch2020decoherence} who found a value of $\Gamma_2^*(\rm P1)= (2 \pi) 16\ \rm kHz/ppm$. The typical concentrations of P1 center go from [REF] for high purity CVD diamond, up to 500 ppm for type 1B HPHT or natural diamonds.

$^{13}$C is much more abundant than P1 center ($\sim$ 10 000 ppm at natural abundance) but being a nuclear spin, it has a gyromagnetic ratio $\approx 2.8\cdot 10^3$ times smaller than that of the P1. This results in a contribution to $\Gamma_2^* \sim 10^3$ times smaller per impurity. It was found experimentally that the broadening caused by $^{13}$C is $\Gamma_2^*(^{13}\rm C ) \approx (2 \pi) 16\ \rm Hz/ppm$ \citep{van1997dependences, barry2020sensitivity}. For natural abundance diamond, this means that $\Gamma_2^*(^{13}\rm C ) \approx (2 \pi) 160\ \rm kHz$.

\subsection{Other spin defects, VH$^-$, War1}
\label{other defects}
Other spin defects are also commonly found in diamond containing NV centers, such as charged vacancy or vacancy clusters \citep{hounsome2006origin}, other nitrogen related defects \citep{newton2007epr}, transition metals \citep{isoya1990fourier}, and especially for CVD grown diamond, hydrogen related defects \citep{hartland2014study}, since the CVD plasma contains up to 95 \% hydrogen.

@@

\subsection{EPR spectroscopy}
Electron paramagnetic resonance (EPR), the electronic equivalent to NMR, is the the most common way to detect electronic spins in solids and liquids. The principle of its detection is the absorption or reflection of a microwave field resonant with a spin transition.

@@

At room temperature, the thermal polarization of a transition at 9500 MHz is $\sim 0.002$. This low value, compared to $\sim 0.8$ for the NV centers, explains in part why EPR is way less sensitive than NV ODMR. The current noise floor for EPR is typically of $\sim 10^{11}$ spins for electronic spins in diamond \citep{mitchell2013x}.

@@

\subsection{Principle of NV center Relaxometry}
Detection \textit{via} cross-relaxation	with NV centers fall under the more general technique of NV relaxometry, where we are effectively measuring a modification in the NV center spin lifetime $T_1$. Relaxometry with NV centers has been used not only to measure resonant cross-relaxation with dark spins \citep{van1989cross, holliday1989optical, epstein2005anisotropic, armstrong2010nv,   hall2016detection, wickenbrock2016microwave,  wood2016wide,  alfasi2019detection, lazda2021cross}, but also to probe the magnetic noise in the diamond environment coming from spins \citep{steinert2013magnetic}, ions\citep{tetienne2013spin}, free radicals \citep{nie2021quantum}, magnetic domains \citep{finco2021imaging}...

>>>>

\documentclass[a4paper, 11pt]{book}

\usepackage{psl-cover}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric-comp,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{bib_thesis.bib} %Import the bibliography file

@@

\title{Cross-relaxation in dense ensembles of NV centers and application to magnetometry}

\author{Clément Pellet-Mary}

\institute{ENS Paris}
\doctoralschool{EDPIF}{564}
\specialty{Physique}
\date{07 Décembre 2022}

\jurymember{1}{Dmitry Budker}{Professor JGU Mainz}{Rapporteur}
\jurymember{2}{Philippe Tamarat}{Professeur, Université de Bordeaux}{Rapporteur}
\jurymember{3}{Maria Chamarro}{Professeure, Sorbonne Université}{Examinateur}
\jurymember{4}{Vincent Jacques}{Dirrecteur de recherche, CNRS}{Examinateur}
\jurymember{5}{Alexandre Tallaire}{Dirrecteur de recherche, CNRS}{Examinateur}
\jurymember{6}{Gabriel Hétet}{Maitre de conférence, ENS Paris}{Directeur de thèse}
% \jurymember{9}{Prénom NOM}{Titre, établissement}{Invité}
% \jurymember{10}{Prénom NOM}{Titre, établissement}{Invité}

\frabstract{
Le centre NV du diamant est un candidat prometteur pour de nombreuses technologies quantiques, que ce soit pour les communications quantiques, le calcul quantique ou la métrologie quantique. Grace aux progrès de synthétisation du diamant, des échantillons contenant une haute concentration en centres NV sont aujourd’hui réalisables. Cette augmentation de la densité de centres NV s’accompagne d’une augmentation des interactions entre ces derniers, ce qui donne lieu à une physique complexe et encore largement inexplorée. De nombreuses applications récentes, telles que l’hyper-polarisation en RMN, la bio-imagerie ou encore la spin-mécanique repose sur des ensembles denses de centres NV. Comprendre et maîtriser les propriétés d’ensemble de ces derniers est un point crucial pour l’établissement de ces technologies.

Cette thèse étudie les échanges de polarisation, par relaxation croisée, entre des ensembles denses de centres NV, ainsi que les interactions entre les centres NV et d’autres impuretés paramagnétiques présentes dans le diamant. Le but de ces travaux est de mieux comprendre les propriétés des ensembles de centres NV, et d’exploiter certaines de ces propriétés pour de nouvelles applications telles que la détection d’impuretés ou la magnétométrie.

Dans un premier temps, nous détaillerons les propriétés des centres NV et nous introduirons des concepts liés à l’interaction dipolaire entre spins et aux relaxations croisées. Nous verrons ensuite comment les relaxations croisées peuvent être utilisées pour détecter optiquement des résonances de spins qui ne sont pas optiquement actifs, en l’occurrence les défauts VH- et War1 dans des diamants CVD. Enfin nous étudierons les relaxations croisées entre différents sous-groupes de centre NV. Nous verrons d’abord leur influence sur la dynamique du spin et leur origine microscopique, puis nous étudierons leur dépendance avec le champ magnétique et leur potentielle utilisation pour la magnétométrie en champs faible.
}

\enabstract{
  The NV center in diamond is a promising candidate for emerging quantum technologies, including quantum communication, quantum computing and quantum metrology. Thanks to the recent progress in diamond synthesis, samples with high NV center concentration can now be created. This increase in the NV density comes with an increase of the interaction between them, which leads to a rich and complex physics. Numerous applications, such as NMR hyperpolarization, bio-imagery or spin mechanics rely on the use of dense NV ensembles. Understanding and mastering the properties of ensembles of NV centers will be a crucial point in the development of these technologies.

This doctoral thesis focuses on cross-relaxation between dense ensembles of NV centers, as well as on the interaction between NV centers and other diamond paramagnetic impurities. The aim of this work is to better understand the properties of dense NV centers ensembles, and to exploit them for new applications such as the detection of impurities and magnetometry.

In the first part, we will detail the properties of single NV centers and introduce notions regarding dipole-dipole spin coupling and cross-relaxations. We will then show how cross-relaxations can be used to optically detect dark spin resonances, in this case the VH- and War1 defects in CVD-grown diamond. Finally, we will present a study of cross-relaxations between subgroups of NV centers. We will first cover the influence of cross-relaxation on the spin dynamics and discuss its microscopic origin. We will then investigate the role of the magnetic field on cross-relaxations as well as their potential use in low field magnetometry.
}

\frkeywords{centre NV, dynamique de spin, magnétométrie}
\enkeywords{NV center, spin dynamics, magnetometry}

\begin{document}

\maketitle{}
\setcounter{tocdepth}{1}

\tableofcontents

\chapter*{Introduction}

Quantum mechanics was first developed at the beginning of the 20th century to explain physical observations such as the blackbody radiation \citep{planck1900theorie} or the photoelectric effect \citep{einstein1905erzeugung}. The quantum theory was subsequently developed in the first half of the 20th century and showed great success, in particular in giving a physical framework for chemistry \citep{pauling1931nature} and leading to a microscopic understanding of matter \citep{kittel1996introduction}. In the second half of the 20th century, technological development based on the new understanding of light and matter have led to what was retroactively called the first quantum revolution \citep{thew2019focus}. Technologies developed in the 50s and 60s include the semiconductor transistor \citep{bardeen1948transistor}, the laser \citep{maiman1960stimulated}, and nuclear magnetic resonance \citep{abragam1961principles} which have since revolutionized the fields of information science, global communications and medical imaging.

A new field of quantum technology has emerged in the last two decades. This ``second quantum revolution" differs from the first one in the usage of individual quantum systems, such as single photons \citep{gleyzes2007quantum}, atoms \citep{neuhauser1980localized} or electrons \citep{peil1999observing}, which have properties that can not be emulated by larger classical systems. These properties include quantum entanglement, superposition or measurement and form the basis of the newly formed quantum information science \citep{nielsen2002quantum, vedral2006introduction, hayashi2006quantum}.

Central to quantum information science is the idea of quantum bits or qubits \citep{schumacher1996sending}, which are the building blocks of quantum information in analogy to the bits of classical information theory. Physical implementation of a qubit could be any quantum system with two well defined quantum states, as long as these states can be initialized, manipulated and readout \citep{divincenzo2000physical}. Popular qubit candidates include superconducting circuits \citep{nakamura1999coherent, orlando1999superconducting}, photons \citep{bennett1992quantum, zhong2020quantum}, quantum dots \citep{veldhorst2014addressable, zajac2018resonantly}, trapped ions \citep{friis2018observation, wright2019benchmarking} and single crystal defects \citep{jelezko2006single, baranov2011silicon, zhong2015optically}.

Defects in large band gap crystals are sometimes referred to as artificial atoms \citep{buluta2011natural}, as the localized wave function of the electrons surrounding the impurity  give rise to discrete energy levels similar to that of an atom. Of all the possible crystals and defects, the most studied defect for quantum applications is the negatively charged nitrogen-vacancy center in diamond \citep{aharonovich2016solid, de2021materials}, often abbreviated to ``NV center". Although this defect had been known and correctly identified almost 50 years ago \citep{davies1976optical}, interest in the NV center really sparked after it was first isolated in 1997 \citep{gruber1997scanning}. The detection of single NV centers proved that NV centers fluoresce brightly \citep{gruber1997scanning}, are photostable \citep{kurtsiefer2000stable}, and that their electronic spin can be polarized and readout optically \citep{jelezko2004observation}, making them an promising solid state qubit.

Since then, single NV centers have been successfully used to create quantum entanglement over several km \citep{hensen2015loophole}, to store quantum information at room temperature for more than one second \citep{maurer2012room}, to detect single proteins via nuclear magnetic resonance \citep{lovchinsky2016nuclear}, and for many other applications in quantum communication \citep{wehner2018quantum}, quantum computing \citep{de2021materials} or quantum sensing \citep{degen2017quantum}.

As the interest in single NV center grew, diamond synthesis improved significantly \citep{achard2020chemical, barry2020sensitivity, edmonds2020generation}  which benefited not only single NV centers but also NV center ensemble. While single NV center are required for most quantum information processes \citep{ladd2010quantum}, several other applications benefit from a greater concentration of defects and are therefore more suited to operate with NV centers ensemble. Four of these applications are pictured in Fig. \ref{shema introduction}.

\begin{figure}[h!]
\centering
\includegraphics[width=.9\textwidth]{Intro et conclusion/Figures/Fig_intro}
\caption{Example of applications with dense ensemble of NV centers. (a) In vivo nanodiomond biomarkers (from \citep{mohan2010vivo}). (b) Quantum sensing with NV ensemble (from \citep{ku2020imaging}). (c) Dynamic nuclear polarization of an external spin bath (from \citep{healey2021polarization}). (d) Spin mechanics with trapped micro diamonds (from \citep{delord2020spin}).}
\label{shema introduction}
\end{figure}

The first application for NV center ensemble is the use of nano or micro diamonds as fluorescent biomarkers \citep{fu2007characterization, mohan2010vivo}. NV centers are good candidates for biomarkers as they emit a bright and stable photoluminesecence, and because diamond nanocrystals present low toxicity \citep{schirhagl2014nitrogen}. Some applications also use the spin degree of freedom of the NV centers, either to perform background free fluorescence microscopy \citep{chapman2013background} or to monitor the angular motion of the nanodiamonds \citep{mcguinness2011quantum, feng2021association}. For most of these applications, NV ensemble provide a better visibility than single NV centers and are therefore preferred.

The second NV ensemble application is for quantum sensing \citep{degen2017quantum}. NV centers are predominantly used as magnetic field sensors \citep{rondin2014magnetometry, barry2020sensitivity}, although they can also detect electric fields \citep{dolde2011electric, michl2019robust}, temperature \citep{acosta2010temperature, kucsko2013nanometre} ,crystal strain \citep{ovartchaiyapong2014dynamic, doherty2014electronic} or rotation \citep{ledbetter2012gyroscopes, ajoy2012stable}. Both single and ensemble of NV centers can be used for sensing: single centers offer unmatched spatial resolution \citep{mitchell2020colloquium} down to a few nm \citep{lovchinsky2016nuclear}, while ensemble provide better sensitivities \citep{wolf2015subpicotesla, barry2020sensitivity}.

The third NV ensemble application is nuclear magnetic resonance (NMR) dynamical nuclear polarization \citep{eills2022spin}. Dynamic nuclear polarization consists in increasing the nuclear polarization of a sample by transferring electron polarization to the nuclei bath \citep{abragam1978principles}. Because the NV electronic spin can be optically polarized at room temperature, micro and nano diamonds containing NV centers are promising candidates as hyperpolarization agents \citep{tetienne2021prospects}. Nuclear polarization of spins inside the diamond lattice was demonstrated with single NV centers \citep{jacques2009dynamic, smeltzer2009robust} and NV ensemble \citep{king2015room, scheuer2016optically, schwartz2018robust}. Polarization transfer to nuclear spins outside of the diamond is an ongoing challenge \citep{eills2022spin, healey2021polarization}, but dense NV ensemble are believed to be necessary to achieve high external polarization \citep{tetienne2021prospects, rizzato2022polarization}.

Finally, the fourth NV ensemble application is the recent field of spin mechanics \citep{perdriat2021spin}. Similarly to optomechanics \citep{aspelmeyer2014cavity}, the goal of spin mechanics is to couple a quantum system to the mechanical degree of freedom of a macroscopic object. Strong coupling between single NV center and an external mechanical oscillators has been observed \citep{rabl2009strong, kolkowitz2012coherent}, but coupling between NV centers and the motion of the diamond has so far only been achieved with dense NV ensemble \citep{delord2020spin, pellet2021magnetic, perdriat2022angle}. Coupling several NV centers to the same mechanical oscillator could also lead to interesting magnetic phase transitions \citep{wei2015magnetic, ma2017proposal}.

Most of the NV ensemble experiments discussed previously work under the assumption that the ensemble is constituted of $N$ independent single NV centers. However, as often in physics, \textit{more is different} \citep{anderson1972more}. Interaction between the NV center themselves or with their environment can lead to significantly different behavior between ensemble and single spins. These difference include a modification of the charge state dynamics \citep{giri2018coupled} and of the spin dynamics \citep{dobrovitski2008decoherence, jarmola2012temperature, mrozek2015longitudinal, choi2017depolarization}, as well as the apparition of exotic phenomena such as time crystal phase \citep{choi2017observation}, Anderson localization \citep{kucsko2018critical} or superradiance \citep{bradac2017room, angerer2018superradiant}.

The work presented in this manuscript focuses on dipolarly coupled dense ensemble of NV centers. More specifically, we look at spin polarization exchange between the NV centers themselves and with their environment. Polarization exchange (flip-flop, spin diffusion, cross-relaxation, ...) in dense ensemble of spins has been abundantly studied in the context of NMR \citep{abragam1978principles} and spintronics \citep{vzutic2004spintronics}. Because NV centers tend to be relatively dilute, the question of spin exchange between between NV themselves \citep{choi2017depolarization} or with their environment \citep{hall2016detection} has only recently been addressed. However, as samples with higher NV concentration become available \citep{acosta2009diamonds, tallaire2020high, shenderova2019synthesis}, these aspects are becoming more and more relevant.

The goal of this present work is both to better understand the physics of dense NV ensemble, which is crucial for many of the NV ensemble applications cited previously, and to exploit some of these many-body properties for new applications.

\bigskip
This manuscript is organized in the following way:

\medskip
The first chapter introduces the theoretical and experimental concepts used in this manuscript. This includes a brief overview of diamond and NV centers fabrication processes, the physics of single NV centers and a presentation of magnetic dipole-dipole interaction and cross-relaxation.

The second chapter covers the observation of cross-relaxation between NV centers and other spin impurities in CVD-grown diamond. We use measurements on the NV centers to identify these impurities and to quantify some of their attributes. Most of the results of this chapter have been published in  \citep{pellet2021optical} and \citep{ngambou2022improving}.

The third chapter focuses on the dipole-dipole mediated spin relaxation within dense ensemble of NV centers. Following the work presented in \citep{choi2017depolarization}, we introduce the concept of NV-flutcuator to explain this phenomenon and present experimental and theoretical results solidifying this hypothesis. The results of this chapter have been partly published in \citep{pellet2022spin} and \citep{pellet2021magnetic}.

Finally, the fourth chapter presents a study on NV-NV dipolar relaxation under low magnetic field. This chapter covers both the rich physics of NV centers under low magnetic field, and a potential application of the low field depolarization as a new magnetometry protocol, as well as comparisons with other NV magnetometry protocols. The results of this chapter have in large part been published in \citep{pellet2022spin}.

\chapter{Spin properties of single and ensemble NV centers}
The aim of this first chapter is to introduce concepts related to Nitrogen-vacancy (NV) centers, to the diamond material, and to the dipole-dipole interaction which are necessary to explain the work presented in the later chapters.

@@

\subsection{Experimental setup}
\label{Sec setup cahp 1}
\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{chapter 1/Figures/shema_exp}
\caption{Confocal microscopy setup detailed in main text}
\label{setup ch1}
\end{figure}
The three most common experimental degrees of freedom to consider for NV experiments are: the optical illumination, the microwave field and the external magnetic field. Other parameters such as the crystal temperature, the electric field or the crystal stress can also be controlled for specific applications.

Fig. \ref{setup ch1} shows a multi-purpose experimental setup which offers a control over these three degree of freedoms. Unless specified otherwise, this is the experimental setup that was used to perform all the measurements shown in this manuscript.

@@

\subsection{NV spin $T_2$}
\label{sec T2 echo}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/Echo}
\caption{Spin-echo protocol. (a) Pulse seqeunce used. (b) Experimental data obtained on sample CVD-pink, fitted with an exponential decay of time $3.9\ \rm{\mu s}$.} 
\label{Echo}
\end{figure}
\subsubsection{Measurement protocol}
As precised earlier, the exact $T_2$ definition depends on the rephasing scheme used. We will focus here on the simplest known rephasing process which is the spin-echo or Hahn-echo sequence.

@@

\subsection{Cross-relaxations with NV centers}
\label{Sec_CR}
\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/Shema_CR}
\caption{Illustration of cross-relaxations between an NV center and an unpolarized spin. Red dots represent the population in each states, and the arrows the various population transfer mechanisms}
\label{CR_shema}
\end{figure}

@@

The focus of chapter 2 will be on the cross-relaxation between NV centers and other spin impurities in CVD-grown diamond, while the focus of chapter 3 and 4 will be on the cross-relaxation between NV centers and other NV centers, as well as their potential application in magnetometry.

\chapter{Cross-relaxations between NV centers and dark spins in CVD-grown diamond}
Crystal defects are promising candidates for quantum information technologies, either through their optical or spin degree of freedom \citep{aharonovich2016solid, atature2018material, bassett2019quantum}. Spectrocopic studies are a necessary step in identifying and characterizing these defects, with electron paramagnetic resonance (EPR) being the spectrocopic tool of choice to measure the spin properties of the defects \citep{newton2007epr}.

In this chapter, we present a method to characterize the spin of diamond impurities through cross-relaxation with NV centers \citep{hall2016detection,pellet2021optical}. This method allows an all-optical detection of non-NV center spin defects, a method that is both cheaper and more sensitive than traditional EPR. We focus in particular on defects in CVD-grown diamonds which have historically been less studied than HPHT diamond, and for which no prior NV cross-relaxation had been reported.

The chapter is organized as follow: we will first detail the most common paramagnetic impurities in the diamond (the so-called ``dark spins") before explaining the proceeding of NV cross-relaxation experiments, and more generally of NV relaxometry. We will then focus on the detection of two particular defects in a CVD diamond, VH$^-$ and War1, and on their characterization (density, spin parameters, temperature dependence). We will finally conclude on the possible improvements and application of this detection method.

The work presented in this chapter has been published in large part in the articles \citep{pellet2021optical, ngambou2022improving}.

\section{Dark spins in diamond}

\begin{figure}[h!]
\centering
\includegraphics[width=.7\textwidth]{chapter 2/Figures/tronche_de_spin}
\caption{Crystalline structure of three spin defects: the NV center, the P1 enter (substituional Nitrogen) and the VH center.} 
\label{repr NV P1 VH}
\end{figure}

We refer by dark spins to paramagnetic impurities for which the spin degree of freedom is not accessible optically. At room temperature, only two defects in diamond do not have a dark spin: the NV center and the more recently found ST1 center \citep{lee2013readout, john2017bright}. Every other spin impurity identified so far is considered to be dark.

In this section we will present the most common dark spins encountered in diamond, which are simply the most common paramagnetic impurities aside from the NV center, and the relation between the diamond growth process and the presence of these impurities. We will also briefly present EPR spectroscopy which is the most common method to characterize these dark spins.

\subsection{P1 centers and $^{13}$C}
The most common spin defects in NV-rich diamond are the substitutional nitrogen centers N$^0_S$ known as the P1 centers (see Fig. \ref{repr NV P1 VH}), which have an electronic spin $1/2$ and a nuclear spin 1, and the $^{13}$C $1/2$ nuclear spin which represents 1.1\% of natural abundance carbon atoms, the remaining carbon atoms being almost exclusively the spinless $^{12}$C isotope. In NV-rich diamonds, these two impurities are the main causesof the NV center dephasing rates $\Gamma_2^*=1/T_2^*$ and $\Gamma_2=1/T_2$  \citep{barry2020sensitivity}.

The dephasing rate caused by a given impurity is directly proportional to the average diagonal dipole-dipole matrix elements ($\Omega_{\rm shift}$ in (eq. \ref{eq. ff df})) between the two spin species \citep{taylor2008high, bauch2020decoherence}. Since the dipole-dipole interaction scales as $1/r^3$, and the distance $r$ between each spin scales as the inverse cubic root of the concentration,  it is expected that the dephasing rate associated with a particular impurity scales linearly with the concentration of the impurity: $\Gamma_2^*(\rm P1) \propto [P1]$.

In the case of P1 centers, this was verified experimentally by looking at controlled sample with varying concentration of nitrogen \citep{bauch2020decoherence}. The authors of the study found a value of $\Gamma_2^*(\rm P1)= (2 \pi) 16\ \rm kHz/ppm$. For the samples studied in this manuscript [REF], the P1 concentration was

$^{13}$C nuclei are typically much more abundant than P1 center ($\sim$ 10 000 ppm without istopic engineering) but being a nuclear spin, it has a gyromagnetic ratio $\approx 2.8\cdot 10^3$ times smaller than that of the P1. As a result, the contribution of a single $^{13}$C nucleus to the NV dephasing rate is $\sim 10^3$ times that of a single P1 center. It was found experimentally that the broadening caused by $^{13}$C is $\Gamma_2^*(^{13}\rm C ) \approx (2 \pi) 16\ \rm Hz/ppm$ \citep{van1997dependences, barry2020sensitivity}. In samples with natural isotopic abundance of carbon, the inhomogeneous dephasing rate due to the $^{13}$C nuclear spins has a value $\Gamma_2^*(^{13}\rm C ) \approx (2 \pi) 160\ \rm kHz$.

\subsection{Other spin defects, VH$^-$, War1}
\label{other defects}
Other spin defects are also commonly found in diamond containing NV centers, such as charged vacancy or vacancy clusters \citep{hounsome2006origin}, other nitrogen related defects \citep{newton2007epr}, transition metals \citep{isoya1990fourier}, and especially for CVD grown diamond, hydrogen related defects \citep{hartland2014study}, since the CVD plasma contains up to 95 \% hydrogen (see sec. \ref{fab HPHT}).

@@

\subsection{EPR spectroscopy}
\begin{figure}[h!]
\centering
\includegraphics[width=.7\textwidth]{chapter 2/Figures/EPR_spectometer.jpg}
\caption{EPR specrometer. Credits: Bruker Corporation} 
\label{photo EPR}
\end{figure}

Electron paramagnetic resonance (EPR), the electronic equivalent to NMR, is the the most common way to detect electronic spins in solids and liquids. The principle of its detection is the absorption or reflection of a microwave field resonant with a spin transition.

@@

The usual EPR apparatus, shown in Fig. \ref{photo EPR}, is a bulky and expensive lab equipment ($\gtrsim$ 1 M€). In contrast, optical detection such as ODMR is technically easier, less expensive, and allows better spatial resolution.

ODMR can also be significantly more sensitive than EPR: single NV centers are routinely observed in ODMR while the the current EPR noise floor for electronic spins in diamond is typically of $\sim 10^{11}$ spins \citep{mitchell2013x}. This difference in sensitivity comes in part from the better spin polarization \footnote{At room temperature, the thermal polarization of a spin with a transition at 9500 MHz is $\sim 0.002$ while the NV center optical polarization can reach 0.8.}, and in part from the more efficient photon detection in the optical range than in the microwave range.

The key idea of this chapter is to allow the optical detection of spin defects that normally do not show ODMR thanks to cross-relaxation with NV centers.

@@

Detection of dark spins via NV center cross-relaxation fall under the more general technique of NV relaxometry, which consists in measuring a modification of the NV center spin lifetime $T_1$. Relaxometry with NV centers has been used not only to measure cross-relaxation with dark spins \citep{van1989cross, holliday1989optical, epstein2005anisotropic, armstrong2010nv,   hall2016detection, wickenbrock2016microwave,  wood2016wide,  alfasi2019detection, lazda2021cross}, but also to probe the magnetic noise in the diamond environment coming from spins \citep{steinert2013magnetic}, ions\citep{tetienne2013spin}, free radicals \citep{nie2021quantum}or magnetic domains \citep{finco2021imaging}.

The protocols used in NV relaxometry measure either directly the spin $T_1$ or indirectly through the NV photoluminescence.
